Sequential rank agreement methods for comparison of ranked lists

### 2018-01-24 latex/biosts3.tex
Updatedd version

\documentclass[12pt,a4paper]{article}
\pdfoutput=1
\usepackage[utf8]{inputenc}
\usepackage[]{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage[]{color}
\usepackage{subcaption}
\usepackage{natbib}
\usepackage{xspace}
\usepackage{calc}
%\usepackage[noae]{Sweave}
\usepackage{authblk}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{enumitem}
%\usepackage{geometry}
%\geometry{a4paper, portrait, margin=1in}
%
% New commands
%
\newcommand{\ie}{\emph{i.e.}\@\xspace}
\newcommand{\eg}{\emph{e.g.}\@\xspace}

\newcommand{\nn}{\nonumber}

\newcommand{\mcomment}[1]{\textcolor{red}{\textbf{#1}}}

@@

\usepackage[amsthm,thmmarks]{ntheorem}
\theoremstyle{plain}
\newtheorem{theorem}{Theorem}
\newtheorem{definition}{Definition}
\newtheorem{proposition}{Proposition}
\newtheorem{corollary}{Corollary}
\newtheorem{lemma}{Lemma}
\newtheorem{assumption}{Assumption}
\newtheorem{example}{Example}
\renewcommand\proofSymbol{\ensuremath{\blacksquare}}

@@

@@

@@

\begin{abstract}
  The comparison of alternative rankings of a set of items is a
  general and common task in applied statistics. Predictor variables
  are ranked according to magnitude of association with an outcome,
  prediction models rank subjects according to the personalized risk
  of an event, and genetic studies rank genes according to their
  difference in gene expression levels. This article constructs
  measures of agreement among two or more ordered lists. We use the
  standard deviation of the ranks to define a measure of agreement
  that both provides an intuitive interpretation and can be applied to
  any number of lists even if some or all are incomplete. The approach
  can identify changepoints in the agreement of the lists, and the
  sequential changes of agreement as a function of the depth of the
  lists can be compared graphically to a permutation based reference
  set or evaluated using analytical asymptotic results.  The
  usefulness of these tools are illustrated using gene rankings, and
  using data from two Danish ovarian cancer studies where we assess
  the within and between agreement of different statistical
  classification methods.
\end{abstract}

Key words: sequential rank agreement, partial ranking, order
statistic, gene rankings, methods comparison, variable selection,
permutation

@@

Ranking of items or results is common in scientific research and
ranked lists occur naturally as the result of many statistical
applications. Regression methods rank predictor variables according to
magnitude of their association with an outcome, prediction models rank
subjects according to their risk of an event, and genetic studies rank
genes according to their difference in gene expression levels across
samples.

When several rankings of the same items are available, a common
research question is to what extent they agree. In particular, is it
possible to identify an optimal rank until which the lists agree on
the items?  A typical situation arises in high-dimensional genomics
studies when several analysis methods are applied to rank a list of
genes according to their association with a phenotype, treatment
effect or other outcome. A measure of agreement of gene rankings
obtained by different methods could often help to identify which genes
are worth to pursue in further experiments.

Several approaches exist to measure the ``distance'' between two
ranked lists. Of these, Kendall's $\tau$ \citep{kendall1948} and
Spearman’s footrule $\rho$ \citep{Spearman1910} are among the most
well-known, but these measures do not distinguish between agreement in
the top versus towards the bottom of the lists and they both provide a
single measure of the overall distance. \citet{Shieh1998} proposed a
weighted version of Kendall's $\tau$ where each pair of rankings can
be assigned different weights, and \citet{Yilmaz2008} proposed the
$\tau_{ap}$ which places higher emphasis on the top of the lists.
Spearman’s footrule uses the ranks of the variables for calculation of
the distance and the use of ranks are also employed in the $M$ measure
of \citet{Bar-Ilan2006} where the reciprocal rank differences are used
to calculate the similarity measure.

@@

Very recently, \citet{hall:schi:2012} proposed a method for comparing
pairwise rankings and derived the asymptotic distribution of the
endpoint where the two ranked lists no longer are in agreement. Their
approach was based on anchoring one of the two lists and subsequently
generating a sequence of 0s and 1s depending on whether the ranks in
the second list was close to the rank from the anchored
list. \citet{samp:verd:2013} followed up on this idea for pairwise
comparison of lists but used penalties based on a truncated geometric
probability instead of a 0-1 process and they evaluated the
distribution of the endpoint of agreement by computational approaches.
The asymptotic distribution in the \citet{hall:schi:2012} paper is
based on letting the number of \emph{lists} increase to infinity which
is a situation that is only relevant in special cases, whereas the
simulation-based null distribution approach of \citet{samp:verd:2013}
is guaranteed to work also for small samples.

In this article we introduce sequential rank agreement for measuring
agreement among (partially) ranked lists.  The general idea is to
define agreement based on the sequence of ranks from the first $d$
items in each list. As agreement metric we adapt the limits of
agreement known from agreement between quantitative variables
\citep{alt:bland:1983,Carstensen2010} but any measure for agreement
could essentially be used. Our proposed approach allows us to
compare multiple lists simultaneously, it provides a dynamic
measure of agreement as a function of the depth in the lists, it places
higher weight on items at the top of the list, it accommodates partially
observed lists of varying lengths, and has a natural interpretation
that directly relates to the ranks.  Graphical illustration of
sequential rank agreement potentially allows us to infer a
changepoint, i.e., a list depth where a substantial change in the agreement of
the lists occur but we also provide asymptotical and randomization-based graphical
tools to compare the observed rank agreement to the expected agreement
found in non-informative data.  In this sense our approach is a
combination and generalization of some of the ideas of
\citet{Carterette2009} and \citet{Boulesteix2009}. The former compares
two rankings based on the distance between them as measured by a
multivariate Gaussian distribution and the latter presents an overview
of approaches for aggregation of ranked lists including bootstrap and
leave-one-out jackknife approaches. We show the asymptotic
distribution of the endpoint of agreement and discuss how to infer the
distribution in small-sample situations using computational
methods. This enables us to make inferences about the endpoint of
agreement even for situations where there is no actual changepoint
present but just a gradual decline in agreement.

The manuscript is organized as follows: In the next section we define
sequential rank agreement for multiple ranked lists and discuss how to
handle incomplete lists. In section 3 we present and discuss
approaches to evaluate the results obtained from sequential rank
agreement. Finally we apply the sequential rank agreement to two
Danish ovarian cancer studies and compare our method to the method of
\citet{hall:schi:2012} in a small sample simulation study before we
discuss the findings along with possible extensions. The approaches
presented in this manuscript are available in the R package
\texttt{SuperRanker} which can be found on CRAN.
% chunk 1

@@

\begin{table}[tb]
  \caption{Example set of ranked lists. (a) shows the ranked lists of
    items for each of three lists, (b) presents the ranks obtained by
    each item in each of the three lists and (c) shows the cumulative
    set of items up to a given depth in the three lists.}
\begin{center}
\footnotesize{
  \begin{subtable}{4.5cm}%
    \caption{}
      \begin{tabular}{cccc}
        \hline\hline % & \multicolumn{3}{c}{List} \\
        Rank & $R^{-1}_1$ & $R^{-1}_2$ & $R^{-1}_3$ \\ \hline
        1 & A & A & B \\
        2 & B & C & A \\
        3 & C & D & E \\
        4 & D & B & C \\
        5 & E & E & D \\ \hline
    \end{tabular}
  \end{subtable}
% \hfill
\hspace{1em}
%%%
  \begin{subtable}{4cm}%
    \caption{}
    \begin{tabular}{cccc} \hline\hline
  % & \multicolumn{3}{c}{List} \\
    Item & $R_1$ & $R_2$ & $R_3$ \\ \hline
    A & 1 & 1 & 2 \\
    B & 2 & 4 & 1 \\
    C & 3 & 2 & 4 \\
    D & 4 & 3 & 5 \\
    E & 5 & 5 & 3 \\ \hline
  \end{tabular}
\end{subtable}
% \hfill
\hspace{1em}
%%%
\begin{subtable}{4cm}%
  \caption{}
\begin{tabular}{cc} \hline\hline
% Depth   &     \\
Depth  &  $S_d$ \\ \hline
1 & $\{$A, B$\}$\\
2 & $\{$A, B, C$\}$ \\
3 & $\{$A, B, C, D, E$\}$ \\
4 & $\{$A, B, C, D, E$\}$ \\
5 & $\{$A, B, C, D, E$\}$ \\ \hline
\end{tabular}
\end{subtable}
}
\end{center}
\label{tab:example}
\end{table}

@@

For an integer $1\le d\le P$ we define the expected set of unique
items found by merging the first $d$ elements across the possible
lists:
\begin{align}
S(d) &= \left\{X_p; \left(\sum_{r\in\Pi} 1\left(r(X_p)\le d\right) Q(r)\right) >0\right\}\label{eq:sumset}
\end{align}
where $1(\cdot)$ denotes the indicator function.
The empirical counterpart is the set of unique items ranked less than or
equal to $d$ in any of the $L$ lists:
\begin{equation}
\widehat{S}_{L}(d) = \left\{X_p; \left(\frac{1}{L}\sum_{l=1}^L 1\left(R_l(X_p)\le d\right)\right)>0\right\},
\end{equation}
which is exemplified in Panel (c) of Table \ref{tab:example}.

We define the \emph{sequential rank agreement} as the weighted
expected agreement of the items found in the set $S(d)$:
\begin{align}
  \textrm{sra}(d) = \frac{1}{|S(d)|}\sum_{p \in S(d)}A(X_p)\label{def:sra}
\end{align}
where $|S(d)|$ is the cardinality of the set $S(d)$.
The empirical counterpart is equivalently given by
\begin{equation}
  \widehat{\textrm{sra}}_L(d) = \frac{\sum_{\{p \in \widehat{S}_{L}(d)\}}(L-1)\widehat{A}_L(X_p)}{(L-1)|\widehat {S}_{L}(d)|}\label{def:sraEst}
\end{equation}

Values of \textrm{sra} close to zero suggests that the lists agree on
the rankings while larger values suggest disagreement.  Just like in
method comparison studies we prefer smaller limits of agreement
because that means that the differences between the rankings is
small. If the ranked lists are identical then the value of sequential
rank agreement will be zero for all position or depths $d$ down the
lists.  The sequential rank agreement is equivalent to the pooled
variance of the items found in $S(d)$, and it can be interpreted as
the average distance of the individual rankings of the lists from the
average ranking for each of the items we have seen until depth $d$.
For an easier interpretation we suggest to plot
$\sqrt{\widehat{\textrm{sra}}_L(d)}$ corresponding to the pooled
standard deviation.

\subsection{Agreement among fully observed lists}
\label{sec:amfol}

The simplest case occurs when all $L$ lists are fully observed, \ie, when
we observe the rank of all $P$ items for all $L$ lists. Fully
observed ranked lists are common and arise, for example, when
different statistical analysis methods are applied to a single dataset
to produce lists of predictors ranked according to their importance,
or if the same analysis method is applied to data from different
populations.

With fully observed lists we can plot the sequential rank agreement
\eqref{def:sra} as a function of depth $d$. An example is seen in the top
panels of Figure~\ref{fig:example1} where four analysis methods were
used to rank 3051 gene expression values measured on 38 tumor mRNA
samples in order to classify between acute lymphoblastic leukemia
(ALL) and acute myeloid leukemia (AML)
\citep{Golub1999}. Preprocessing of the gene expression data was done
as described in \citet{Dudoit2002} and the four different analysis
approaches were: marginal unequal variances two-sample $t$ tests,
marginal logistic regression analyses, elastic net logistic regression
\citep{friedman2010regularization}, and marginal maximum information
content correlations (MIC) \citep{Reshef2011}. For the first two
methods, the genes were ranked according to minimum $p$ value, for
logistic regression the genes were ordered by size of the
corresponding coefficients (after standardization), and MIC was
ordered by absolute correlation which resulted in the top rankings
seen in Table~\ref{tab1}.

The sequential rank agreement curve shown in the two left plots of
Figure~\ref{fig:example1} roughly indicates the average distance in
ranks for the genes found among the first $d$ positions in the $4$
lists.  It can be seen that the sequential rank agreement limits are
better towards the top of the lists (smaller values on the $y$ axis
corresponds to \emph{better} agreement) than towards the bottom of the
lists. It is clear from the curve in Figure~\ref{fig:example1} that
there is a substantial drop in agreement (higher sra) even after the
first depth. Thus, if we were to restrict attention to a small set of
predictors then we would focus on predictors 2124, 829, and 378 as
seen in Table~\ref{tab1}.

% latex table generated in R 3.2.3 by xtable 1.8-2 package
% Fri Apr  8 10:22:58 2016
\begin{table}[tb]
\centering
\caption{List of ranked results from the Golub data. Numbers indicate the predictor/gene for the given ranking and method. Only the top 10 ranks are shown in the table.} 
\label{tab1}
\begin{tabular}{rrrrr}
  \hline
Ranking & Welsh's $t$ & LogReg & ElasticNet & MIC \\ 
  \hline
1 & 2124 & 2124 & 829 & 378 \\ 
  2 & 896 & 896 & 2124 & 829 \\ 
  3 & 2600 & 829 & 2198 & 896 \\ 
  4 & 766 & 394 & 1665 & 1037 \\ 
  5 & 829 & 766 & 1920 & 2124 \\ 
  6 & 2851 & 2670 & 1042 & 808 \\ 
  7 & 703 & 2939 & 808 & 108 \\ 
  8 & 2386 & 2386 & 849 & 515 \\ 
  9 & 2645 & 1834 & 937 & 2670 \\ 
  10 & 2002 & 378 & 1995 & 2600 \\ 
   \hline
\end{tabular}
\end{table}

@@

\begin{figure}[htbp]
   \begin{center}
 \includegraphics[width=.49\textwidth]{paper-fig1a}
 \includegraphics[width=.49\textwidth]{paper-fig2a}
 \includegraphics[width=.49\textwidth]{paper-fig1b}
 \includegraphics[width=.49\textwidth]{paper-fig2b}
 \end{center}
 \caption{Left panels: Sequential rank agreement for 4 different
   analysis methods applied to the 3051 genes in the Golub data (black
   line). Right panels: Corresponding sequential rank agreement for
   the same data but where only the top 20 ranked items are available
   and the rank of the remaining items are not available. The blue and
   red areas correspond to the independent and randomized reference
   hypothesis areas, respectively. The bottom plots are identical to
   the corresponding top plots but have been zoomed in on the first
   part of the $x$ axis.}
 \label{fig:example1}
 \end{figure}

% If the lists agree on all items then the sequential rank agreement is
% zero for all depths. 
 Generally, if there are changes in the level of
 rank agreement then this suggests that there are sets of items that
 are ranked similarly in all lists and other sets of items that have
 been assigned vastly different ranks in the lists. For example, in gene
 association studies we generally expect the agreement among the lists
 to be better towards the top of the lists and worse towards the end
 of the lists. In this case the sequential rank agreement curve will
 start at a low level and then increase until it levels off as seen in
 Figure~\ref{fig:example1}.

\subsection{Analysis of incomplete lists}
Incomplete or partial lists are a common occurrence that arise, for
example, in case of missing data (items), when comparing top $d$ list
results from publications, or when some methods only rank a subset of
the items. For example, penalized regression based on the Lasso
provides a sparse set of predictors that have non-zero
coefficients. There is no obvious ordering of the set of predictors
whose coefficient has been shrunk to zero and thus we end up with a
partial ordering.  Incomplete lists also occur if the analyst
restricts attention to the ranks of items that have been found to be
statistically significant.

@@

When the two assumptions are satisfied then it is clear that the
interesting part of the sequential rank agreement curves is restricted
to the depths where the number of items without ranks available is
low. Generally, without additional prior knowledge about the
underlying distribution of the ranks it seems reasonable to restrict
attention of the sequential rank agreement to lists no longer than the
longest of the $L$ lists examined.

Like for fully observed lists we generally expect the sequential rank
agreement to start low and then increase unless the lists are
completely unrelated (in which case the sequential rank agreement will
be constant at a high level) or if the lists mostly agree on the
ranking (in which case the sequential rank agreement will also be
constant but at a low level). For incomplete ranked lists we also
expect a changepoint around the depth where the lists are
become incomplete. This is an artefact stemming from the fact that we assume
that the remainder of the lists can be replaced by a simple
permutation of the missing items.

The two right-most plots of Figure~\ref{fig:example1} show the impact
of restricting the Golub data such that only top-20 lists are
available instead of full lists of length 1000 (20 was chosen to
resemble the list lengths that might be published in a
manuscript). The sequential rank agreement increases much quicker
because the incomplete lists introduce more noise in the estimation of
the agreement, but it is still possible to see that the top of the
list has a sequantial rank agreement that is not substantially
different from the full lists.

@@

\section{Evaluating sequential rank agreement}
To interpret the sequential rank agreement values we propose two
different benchmark values corresponding to two different
hypotheses. We wish to determine if we observe better agreement than
what would be expected if there were no relevant information available
in the data.

@@

Alternatively, we can remove the restriction on the independence among
the methods used to generate the individuals rankings in the $L$ lists
by only requiring that there is no information contained in the ranking but that
the rankings are all based on applying the method/approaches to the
same data
\begin{eqnarray*}
\widetilde H_0 & :&  \text{The list rankings are based on data that contain}\\
& &   \text{no association to the outcome.}
\end{eqnarray*}

@@

$H_0$ is a quite unrealistic null hypothesis but we can easily obtain
realizations from that null hypothesis simply by permuting the items
within each list and then computing the sequential rank agreement for
the permuted lists. In the fully observed case each experiment
contains $L$ lists of random permutations of the items in $X$. For the
incomplete case we first permute the items $X_1,\dots,X_P$ and then assign missing ranks
for list $l$ from  $d_l$ to $P$ (\ie, each list so it has the same number of
observed rankings as was observed for list $l$ in the original
dataset). The sequential rank agreement curve from the original
lists can then be compared to, say, the pointwise 95\% quantiles of
the observed rank agreements obtained under $H_0$.

>>>>

\documentclass[oupdraft]{bio}
%\usepackage[colorlinks=true, urlcolor=citecolor, linkcolor=citecolor, citecolor=citecolor]{hyperref}
% \documentclass[12pt,a4paper]{article}
\pdfoutput=1
\usepackage[utf8]{inputenc}
\usepackage[]{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage[]{color}
\usepackage{subcaption}
% \usepackage{natbib}
\usepackage{xspace}
\usepackage{calc}
% \usepackage[noae]{Sweave}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{xcolor}

% \usepackage{authblk}

\usepackage[nolists,nomarkers]{endfloat}

%\usepackage{geometry}
%\geometry{a4paper, portrait, margin=1in}
%
% New commands
%
\newcommand{\ie}{\emph{i.e.}\@\xspace}
\newcommand{\eg}{\emph{e.g.}\@\xspace}
\newcommand{\nn}{\nonumber}

\providecolor{added}{rgb}{1,0,0}
\newcommand{\added}[1]{{\color{added}{}#1}}
% \newcommand{\added}[1]{#1}   % Use after correction

% \newcommand{\mcomment}[1]{\textcolor{red}{\textbf{#1}}}

@@

%\usepackage[amsthm,thmmarks]{ntheorem}
% \theoremstyle{plain}
% \newtheorem{theorem}{Theorem}
%\newtheorem{definition}{Definition}
%\newtheorem{proposition}{Proposition}
\newtheorem{corollary}{\normalfont\scshape Corollary}[section]
%\newtheorem{lemma}{Lemma}
%\newtheorem{assumption}{Assumption}
%\newtheorem{example}{Example}
%\renewcommand\proofSymbol{\ensuremath{\blacksquare}}

@@

\title{Sequential rank agreement methods for comparison of ranked lists}

\author{CLAUS THORN EKSTRØM$^\ast$, THOMAS ALEXANDER GERDS, ANDREAS
  KRYGER JENSEN\\[4pt]
%
\textit{%
Biostatistics, Department of Public Health,
University of Copenhagen,
Øster Farimagsgade 5 B, P.O.B. 2099,
DK-1014 Copenhagen K, Denmark}
\\[2pt]
% E-mail address for correspondence
{ekstrom@sund.ku.dk}}

@@

% Running headers of paper:
\markboth%
% First field is the short list of authors
{C. T. Ekstrøm and others}
% Second field is the short title of the paper
{Sequential rank agreement for comparison of ranked
  lists}

@@

\footnotetext{To whom correspondence should be addressed.}

\begin{abstract} { The comparison of alternative rankings of a set of
    items is a general and common task in applied
    statistics. Predictor variables are ranked according to magnitude
    of association with an outcome, prediction models rank subjects
    according to the personalized risk of an event, and genetic
    studies rank genes according to their difference in gene
    expression levels. \added{We propose a sequential rank agreement
      measure to quantify the rank agreement among two or more ordered
      lists. This measure has an intuitive interpretation, it can be
      applied to any number of lists even if some are partially
      incomplete, and it provides information about the agreement
      along the lists. The sequential rank agreement can be evaluated
      analytically or be compared graphically to a permutation based
      reference set in order to identify changes in the list
      agreements.}  The usefulness of this measure is illustrated
    using gene rankings, and using data from two Danish ovarian cancer
    studies where we assess the within and between agreement of
    different statistical classification methods.}  {rank agreement;
    incomplete ranking; order statistic; methods comparison;
    permutation}
\end{abstract}

%Key words: rank agreement, partial ranking, order
%statistic, gene rankings, methods comparison, variable selection,
%permutation

@@

Ranking of items or results is common in scientific research and
ranked lists occur naturally as the result of many statistical
applications. Regression methods rank predictor variables according to
magnitude of their association with an outcome, prediction models rank
subjects according to their risk of an event, and genetic studies rank
genes according to their difference in gene expression levels across
samples.  \added{Two common research questions are of interest when
  several rankings of the same items are available: 1) To what extent
  do the lists agree on the rankings and how will that change as we go
  through the lists, and 2) Is it possible to identify an optimal rank
  until which the lists agree on the items?

  A typical situation where
  these questions arise is in high-dimensional genomics studies such
  as genome-wide association studies where several analysis methods
  (e.g., regression methods, lasso, random forest) can be used
  identify and rank millions of gene variants according to their
  association with the outcome. The importance (i.e., ranking) of each
  gene variant may vary from method to method and a consensus summary
  of agreement of the findings is needed to determine which gene
  variants to investigate more closely in subsequent validation
  studies. To minimiize expenses it is only of interest to consider
  gene variants that have high ranking across the different methods.
  Another situation where multiple ranked lists appear is in machine
  learning where the stability of the ranks produced by a
  ``black-box''-technique can be evaluted by bootstrapping the data
  repeatedly and comparing the ranks obtained from training the models
  on the bootstrapped data. Assessing which items are stable (i.e.,
  consistant across bootstrap samples) will help to weed out possible
  fluke findings.}

In this article we introduce sequential rank agreement for measuring
agreement among  ranked lists.  The general idea is to
define agreement based on the sequence of ranks from a subset of the first $d$
items in each list. As agreement metric we adapt the limits of
agreement known from agreement between quantitative variables
\citep{alt:bland:1983,Carstensen2010} but any measure for agreement
could essentially be used. Our proposed approach allows us to
compare multiple lists simultaneously, it provides a dynamic
measure of agreement as a function of the depth in the lists, it places
higher weight on items at the top of the list, it accommodates partially
observed lists of varying lengths, and has a natural interpretation
that directly relates to the ranks.  Graphical illustration of
sequential rank agreement allows us to infer any
changepoints, i.e., a list depth where a substantial change in the agreement of
the lists occur but we also provide asymptotical and randomization-based graphical
tools to compare the observed rank agreement to the expected agreement
found in non-informative data.  In this sense our approach is a
combination and generalization of some of the ideas of
\citet{Carterette2009} and \citet{Boulesteix2009}. The former compares
two rankings based on the distance between them as measured by a
multivariate Gaussian distribution and the latter presents an overview
of approaches for aggregation of ranked lists including bootstrap and
leave-one-out jackknife approaches. We show the asymptotic
distribution of the endpoint of agreement and discuss how to infer the
distribution in small-sample situations using computational
methods. This enables us to make inferences about the endpoint of
agreement even for situations where there is no actual changepoint
present but just a gradual decline in agreement.

% Several approaches exist to measure the ``distance'' between two
% ranked lists. Of these, Kendall's $\tau$ \citep{kendall1948} and
% Spearman’s footrule $\rho$ \citep{Spearman1910} are among the most
% well-known, but these measures do not distinguish between agreement in
% the top versus towards the bottom of the lists and they both provide a
% single measure of the overall distance. \citet{Shieh1998} proposed a
% weighted version of Kendall's $\tau$ where each pair of rankings can
% be assigned different weights, and \citet{Yilmaz2008} proposed the
% $\tau_{ap}$ which places higher emphasis on the top of the lists.
% Spearman’s footrule uses the ranks of the variables for calculation of
% the distance and the use of ranks are also employed in the $M$ measure
% of \citet{Bar-Ilan2006} where the reciprocal rank differences are used
% to calculate the similarity measure.

@@

Very recently, \citet{hall:schi:2012} proposed a method for comparing
pairwise rankings and derived the asymptotic distribution of the
endpoint where the two ranked lists no longer are in agreement. Their
approach was based on anchoring one of the two lists and subsequently
generating a sequence of 0s and 1s depending on whether the ranks in
the second list was close to the rank from the anchored
list. \citet{samp:verd:2013} followed up on this idea for pairwise
comparison of lists but used penalties based on a truncated geometric
probability instead of a 0-1 process and they evaluated the
distribution of the endpoint of agreement by computational approaches.
The asymptotic distribution in the \citet{hall:schi:2012} paper is
based on letting the number of \emph{lists} increase to infinity which
is a situation that is only relevant in special cases, whereas the
\added{simulation-based null distribution approach of
  \citet{samp:verd:2013} does not rely on asymptotic results to
  evaluate their pairwise findings.}

The manuscript is organized as follows: The next section defines
sequential rank agreement for multiple ranked lists and discuss how to
handle incomplete lists. In section 3 we discuss approaches to
evaluate the results obtained from sequential rank agreement. Finally
we apply the sequential rank agreement to two Danish ovarian cancer
studies and compare our method to the method of \citet{hall:schi:2012}
in a small sample simulation study before we discuss the findings
along with possible extensions. The approaches presented in this
manuscript are available in the R package \texttt{SuperRanker} which
can be found on CRAN.
% chunk 1

@@

@@

For an integer $1\le d\le P$ we define the expected set of unique
items found by merging the first $d$ elements across the possible
lists:
\begin{align}
S(d) &= \left\{X_p; \left(\sum_{r\in\Pi} 1\left(r(X_p)\le d\right) Q(r)\right) > \added{\varepsilon}\right\}\label{eq:sumset}
\end{align}
where $1(\cdot)$ denotes the indicator function\added{, and where $\varepsilon\in[0,1)$ is a pre-specified constant that sets the minimum proportion of lists that an item must be present in before it is added to $S(d)$. When $\varepsilon=0$ then an item is included as soon as it is present in just one list.}
The empirical counterpart is the set of unique items ranked less than or
equal to $d$ in any of the $L$ lists:
\begin{equation}
\widehat{S}_{L}(d) = \left\{X_p; \left(\frac{1}{L}\sum_{l=1}^L 1\left(R_l(X_p)\le d\right)\right)>\added{\varepsilon}\right\},
\end{equation}
which is exemplified in Panel (c) of Table \ref{tab:example}.

We define the \emph{sequential rank agreement} as the weighted
expected agreement of the items found in the set $S(d)$:
\begin{equation}
\added{  \textrm{sra}(d) = \left\{\begin{array}{cl}\frac{1}{|S(d)|}\sum_{p \in S(d)}A(X_p)  & \text{when $|S(d)|>0$} \\ 0 & \text{otherwise}\end{array}  \right.}\label{def:sra}
\end{equation}
where $|S(d)|$ is the cardinality of the set $S(d)$. \added{As stated, we are only interested in sra($d$) when $|S(d)|>0$ so the value 0 above is somewhat arbitrary.}
The empirical counterpart when $|S(d)|>0$ is equivalently given by
\begin{equation}
  \widehat{\textrm{sra}}_L(d) = \frac{\sum_{\{p \in \widehat{S}_{L}(d)\}}(L-1)\widehat{A}_L(X_p)}{(L-1)|\widehat {S}_{L}(d)|}.\label{def:sraEst}
\end{equation}

Values of \textrm{sra} close to zero when $|S(d)|>0$ suggest that the
lists agree on the rankings while larger values suggest
disagreement. \added{If $|S(d)|=0$ then no items were sufficiently
  frequent among the observed lists and we can conclude that the lists
  do not agree above the threshold $\varepsilon$.  } If the ranked
lists are identical then the value of sequential rank agreement will
be zero for all values of $d$. %position or depths $d$ down the lists.

\subsection{Interpreting and applying sequential rank agreement}
\label{sec:amfol}

The sequential rank agreement is equivalent to the pooled variance of
the items found in $S(d)$. Thus, the square root of the sequential
rank agreement measures the average of the average difference in
rankings among the lists for the items we have included until depth
$d$.

In method comparison studies the observed agreement is compared to a
pre-specified acceptable limit and we can do similarly. For easy
visualization of the rank agreement we suggest to plot
$\sqrt{\widehat{\textrm{sra}}_L(d)}$ corresponding to the pooled SD
against $d$.

As an example consider the data by \citet{Golub1999} (also found in
\citet{Dudoit2002}) where 3051 gene expression values measured on 38
tumor mRNA samples were used to classify between acute lymphoblastic
leukemia (ALL) and acute myeloid leukemia (AML). Several possible
analysis methods can be used for this for example marginal unequal
variances two-sample $t$ tests, marginal logistic regression analyses,
elastic net logistic regression \citep{friedman2010regularization},
and marginal maximum information content correlations (MIC)
\citep{Reshef2011}, and we would like to identify a set of genes that
consistently are most likely to be associated to leukemia.

\added{
For the first two methods, the genes were ranked according to minimum
$p$ value, for logistic regression the genes were ordered by size of
the corresponding coefficients (after standardization), and MIC was
ordered by absolute correlation which resulted in the top rankings
seen in Table~\ref{tab1}. The ranked lists appear to agree that genes
2124 and 829 are among the most interesting while the best ranked
gene from MIC, gene 378, is not found in the top 10 for two of the
other methods.

% The simplest case occurs when all $L$ lists are fully observed, \ie, when
% we observe the rank of all $P$ items for all $L$ lists. Fully
% observed ranked lists are common and arise, for example, when
% different statistical methods are applied to a dataset
% to produce lists of predictors ranked according to their importance,
% or if the same analysis method is applied to data from different
% populations.

%With fully observed lists we can plot the sequential rank agreement
%\eqref{def:sra} as a function of depth $d$. An example is seen in the top
%panels of Figure~\ref{fig:example1} where four analysis methods were
%used to rank 3051 gene expression values measured on 38 tumor mRNA
%samples in order to classify between acute lymphoblastic leukemia
%(ALL) and acute myeloid leukemia (AML)
%\citep{Golub1999}. Preprocessing of the gene expression data was done
% as described in \citet{Dudoit2002} and the four different analysis
% approaches were:

The sequential rank agreement curve (using $\varepsilon=0$ to decide
that an item should be present in just a single list before it is
included) shown in the left plot of Figure~\ref{fig:example1} show the
average distance in ranks for the genes considered among the first $d$
positions. Not surprisingly, the sequential rank agreement is better
towards the top of the lists (smaller values on the $y$ axis
corresponds to \emph{better} agreement) than towards the bottom of the
lists. It is clear from the curve in Figure~\ref{fig:example1} that
there is a substantial deterioration in agreement (higher sra) after
depth 5. Thus, if we were to restrict attention to a small set of
predictors then our prime focus would be on the items found among the
top-5 lists from Table~\ref{tab1}). The choice of which depth until
which we think the lists agree can be chosen either from a
pre-specified threshold for an acceptable difference in rankings or
from a pre-specified item set size. A changepoint analysis on the
sequential rank agreement would be able to identify depths where a
substantial increase/change in rank agreement occurs and would be
another way to identify sets of items that share agreement among the
lists if a pre-specified acceptable rank agreement threshold is not
given.}

% latex table generated in R 3.2.3 by xtable 1.8-2 package
% Fri Apr  8 10:22:58 2016

@@

% If the lists agree on all items then the sequential rank agreement is
% zero for all depths. 
Generally, changes in the level of rank agreement suggest that there
are sets of items that are ranked similarly in all lists while other
items constitute set(s) that have been assigned vastly different
ranks. When the lists are likely to agree on a few top ranked items
then the sequential rank agreement curve will start at a low level and
then increase until it levels off exactly as is seen in
Figure~\ref{fig:example1}.

\subsection{Analysis of incomplete lists}
Incomplete or partial lists are a common occurrence that arise, for
example, in case of missing data (items), when comparing top $d$ list
results from publications, or when some methods only rank a subset of
the items. For example, penalized regression such as the Lasso
provides a sparse set of predictors that have non-zero
coefficients. There is no obvious ordering of the set of predictors
whose coefficient has been shrunk to zero and thus we end up with a
partial ordering.  Incomplete lists also occur if for example the analyst
restricts attention to the ranks of items that have been found to be
statistically significant.

@@

When the two assumptions are satisfied then it is clear that the
interesting part of the sequential rank agreement curves is restricted
to depths where the number of items without ranks available is
low.
%Generally, without additional prior knowledge about the
%underlying distribution of the ranks it seems reasonable to restrict
%attention of the sequential rank agreement to lists no longer than the
%longest of the $L$ lists examined.
Similar to fully observed lists we generally expect the sequential
rank agreement to start low and then increase unless the lists are
completely unrelated (in which case the sequential rank agreement will
be constant at a high level) or if the lists mostly agree on the
ranking (in which case the sequential rank agreement will also be
constant but at a low level). For incomplete ranked lists we also
expect a changepoint around the depth where the lists are become
incomplete. This is an artefact stemming from the fact that we assume
that the remainder of the lists can be replaced by a simple
permutation of the missing items. \added{Note that if an item is
  ranked highly in a few lists but unranked in the remaining lists
  then it gets poor rank agreement since we only compare
  whether the lists agree on their rankings and in this case they
  clearly do not.}

The right-most plot of Figure~\ref{fig:example1} shows the impact
of restricting the Golub data such that only top-20 lists are
available instead of full lists of length 3051 (20 was chosen to
resemble the list lengths that might be harvested from published manuscripts). The sequential rank agreement increases much quicker
because the incomplete lists introduce more noise in the estimation of
the agreement, but it is still possible to see that the top of the
list has a sequantial rank agreement that is not substantially
different from the full lists.

@@

\section{Evaluating sequential rank agreement results}
To evaluate the sequential rank agreement values we propose two
different benchmark values corresponding to two different
hypotheses. We wish to determine if we observe better agreement than
would be expected if there were no relevant information available in
the data.

@@

Alternatively, we can remove the restriction of the independence among
the methods used to generate the $L$ ranked lists and only require
that there is no information contained in the rankings but that the
rankings are all based on applying the method/approaches to the same
data
\begin{eqnarray*}
\widetilde H_0 & :&  \text{The list rankings are based on data that contain}\\
& &   \text{no association to the outcome.}
\end{eqnarray*}

@@

$H_0$ is a quite unrealistic null hypothesis but we can easily obtain
realizations from that null hypothesis simply by permuting the items
within each list and then computing the sequential rank agreement for
the permuted lists. In the fully observed case each experiment
contains $L$ lists of random permutations of the items in $X$. For the
incomplete case we first permute the items $X_1,\dots,X_P$ and then assign missing ranks
for list $l$ from  $d_l$ to $P$ (\ie, each list has the same number of
observed rankings as was observed for list $l$ in the original
dataset). The sequential rank agreement curve from the original
lists can then be compared to, say, the pointwise 95\% quantiles of
the observed rank agreements obtained under $H_0$.
